---
title: "Logistic Regression 1 & 2"
format: 
  clean-typst:
    include-in-header: "custom.typ"
    commands: [small-cite]
execute: 
  echo: false
  warning: false    
author:
  - name: A Pathmeswaran
    orcid: 0000-0003-4065-2639
    email: pathmes@kln.ac.lk
date: 2026-03-12
date-format: long
handout: false
brand:
  typography: 
    fonts: 
      - family: Roboto
        weight: [100, 200, 300, 400, 500, 600, 700]
        style: [normal, italic]
        source: google
    base: Roboto
    headings: Roboto
---


```{r}
#| label: loadPacks

library(tidyverse)
library(aod)
library(summarytools)
library(gt)
library(gtsummary)

```

```{r}
#| label: getData

source(here::here("kirkwood19.2.R"))
```


### Syllabus
- Logistic regression model
- interpretation of coefficients for dichotomous & continuous variables

 . . .  
 
- Multiple logistic regression:  
  a. Wald test
  b. Likelihood ratio test
  c. Dummy variables
  d. Model building strategies
  e. Goodness of fit tests
  
### Logit

- Logit is equal to log of odds  
- If p is a proportion  

$$logit(p) = log_e (\frac{p}{1-p})$$

### Likelihood

- Probability
  - What is the probability of the observed data for a stated probability
  - What is the probability of of observing 5 heads when tossing a fair coin five times
  
- Likelihood
  - What is the likelihood of a particular hypothesis to result in the observed data
  - What is the likelihood of a bag to have 80% red balls when 8 out of 10 balls drawn frommthe bag are red.

### Data from Kirkwoods book

```{r}
set.seed(21022025)
t3[sample(nrow(t3), 7), ] |> 
  kableExtra::kable(row.names = FALSE)
```

### MF postivity by area

```{r}
ctable(x = t3$Area,
         y = t3$Microfilaria,
       totals = F)
```
```{r}
tx <- table(t3$Area, t3$Microfilaria)
osv <- tx[1,2]/tx[1,1]
orf <- tx[2,2]/tx[2,1]
```

 . . .  
 
Odds of MF positivity in Rainforest - `r tx[2,2]` / `r tx[2,1]` = 
`r round(orf, 2)`  
Odds of MF positivity in Savannah - `r tx[1,2]` / `r tx[1,1]` = 
`r round(osv, 2)`


### Table with OR

```{r}
or_fun <- function(data, variable, by, ...) {
  table(data[[by]], data[[variable]]) %>%
    fisher.test() %>%
    broom::tidy()
}

  t3 %>%
  select(Microfilaria, Area) %>%
  tbl_summary(by = Area, missing = "no", type = all_dichotomous() ~ "categorical") %>%
  add_difference(test = Microfilaria ~ or_fun, estimate_fun = Microfilaria ~ style_ratio) %>%
  modify_header(estimate ~ "**Odds Ratio**") |> 
  as_gt() |> 
tab_options(table.font.size = 24)
```

### Properties / Advantages of Odds Ratios

1. When the outcome is rare the odds ratio is the same as the risk ratio
2. The conclusions are identical whether the outcome is the occurence of an event or the absence of the event.
2. Unlike the risk ratio the odds ratio is not constrained when the outcome is common in the unexposed group.

### MF postivity by area 2

```{r}
ctable(x = t3$Area,
         y = t3$Microfilaria,
         OR = TRUE,
         chisq = TRUE)
```

### Logistic regression{.smaller}

#### Odds Ratio in Case Control Studies
$$\text{Odds Ratio} = \frac{\text{Odds of Exposure among Cases}}
{\text{Odds of Exposure among Controls}}$$
Same as  
$$\text{Exposure Odds Ratio} = \frac{\text{Odds in Exposed Group}}
{\text{Odds in Unexposed Group}}$$

$$\text{Odds in Exposed Group} = \text{Odds in Unexposed Group} \times \text{Exposure Odds Ratio}$$

## Logistic regression output




### Raw

```{r}
m1 <- glm(
  Microfilaria ~ Area, 
  data = t3, 
  family = "binomial")
summary(m1)
```

### OR

```{r}
exp(cbind(OR = coef(m1), confint(m1)))
```

### Print 

```{r}
gtsummary::tbl_regression(m1)|> 
  as_gt()|> 
tab_options(table.font.size = 24) 
```


```{r}
gtsummary::tbl_regression(m1, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 24)
```





### Birth Weight Data

```{r}
load("C:/AP/MyR/lectures/PGIM2/MD_CM_2023/md2023stats/bwt.RData")
bwt$lbw <- bwt$bw < 2500
bwt$ht_cm <- bwt$height*100
bwt$ht_dm <- bwt$height*10

bwt |> select(height, bw, lbw) |> 
  tbl_summary(
     type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{median} ({p25}, {p75})", "{mean}, ({sd})")
  ) |> 
  as_gt() |> 
tab_options(table.font.size = 24)
```

### Maternal height and LBW

```{r}
m2 <- glm(
  lbw ~ ht_cm, 
  data = bwt, 
  family = "binomial")

gtsummary::tbl_regression(m2, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 24)


```

### Maternal height and LBW (raw)

```{r}
m2 <- glm(
  lbw ~ ht_cm, 
  data = bwt, 
  family = "binomial")

int <- as.numeric(m2$coefficients[1])
beta <- as.numeric(m2$coefficients[2])

m2|> summary.glm() |> coef() |> round(3)
cbind(beta = coef(m2), confint(m2)) |> round(3)
exp(cbind(OR = coef(m2), confint(m2))) |> round(2)
```

### Maternal height (in dm) and LBW

```{r}
m3 <- glm(
  lbw ~ ht_dm, 
  data = bwt, 
  family = "binomial")

gtsummary::tbl_regression(m3, exponentiate = FALSE)|> 
  as_gt() |> 
tab_options(table.font.size = 32)

cbind(Beta = coef(m3), confint(m3))
```

### LBW Predicted Probability 

$$p = \frac{odds}{(1 + odds)}$$
$$Odds_\text{ Exposed Group} = Odds_\text{ Unexposed Group} \times OR$$

Calculate the probability of LBW for mother's height = 140

### Predicted Odds

$$Odds_\text{ Exposed Group} = Odds_\text{ Unexposed Group} \times OR$$

$$Odds_\text{(Maternal Height = 140)} = Odds_\text{ Unexposed Group} \times OR_{Height}^{ 140}$$


$$Odds_\text{(Maternal Height = 140)} = e^{(Intercept + \beta_{Height} \times 140)}$$
```{r}
#| echo: true

oddsH140 <- exp(int + beta*140)
print(paste("Maternal Height 140 cm - Odds of LBW =", round(oddsH140, 3)))
Pr140 <- oddsH140/ (1 + oddsH140)
print(paste("Maternal Height 140 cm - Probability of LBW =", round(Pr140, 2)))
```




### LBW Predicted Probability for selected maternal heights

```{r}
h <- c(round(mean(bwt$ht_cm),1), seq(140, 180, 10))
pt <- cbind(h, exp(predict(m2, data.frame(ht_cm = h))),
      predict(m2, data.frame(ht_cm = h), type = "response"))

colnames(pt) <- c("Height", "P_odds", "P_pob")
as_tibble(pt) |>
  kableExtra::kable(digits = c(1, 3, 3))
```
### What if we had used a linear model
```{r}
mg <- glm(
lbw ~ ht_cm,
data = bwt,
family = "gaussian")

gtsummary::tbl_regression(mg, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 24)

cbind(beta = coef(mg), confint(mg))


```

### LBW Predicted Probability (based on linear regression) for selected maternal heights

 . . .  
 
 {{< v 1em >}}
 
 
```{r}
cbind(h, predict(mg, data.frame(ht_cm = h)))|> 
  kableExtra::kable(row.names = FALSE, digits = 2,
                    col.names = c("height", "predicted pr")) 
```
### Birthweight by Maternal Education

```{r}
bwt$edu <- as.character(bwt$education)
bwt$edu3 <- case_match(
bwt$edu,
c("TEC", "GDU", "PG", "1-5", "6-8") ~ "e1",
c("8-10", "OL") ~ "e2",
c("AL", "NONE" ) ~ "e3"
)

bwt$edu3[bwt$idn == 54] <- "e3"
bwt$edu3[bwt$idn == 66] <- "e3"
bwt$edu3[bwt$idn == 182] <- "e3"
bwt$edu3[bwt$idn == 239] <- "e3"
bwt$edu3[bwt$idn == 259] <- "e3"
bwt$edu3[bwt$idn == 261] <- "e3"
bwt$edu3[bwt$idn == 432] <- "e3"

m5 <- glm(
lbw ~ edu3,
data = bwt,
family = "binomial")

bwt |> 
  select(lbw, edu3)  |> 
  tbl_summary(by = edu3, missing = "no", type = all_dichotomous() ~ "categorical") |> 
as_gt() |> 
tab_options(table.font.size = 24)

```
### LBW by Education - Logistic Regression
```{r}
gtsummary::tbl_regression(m5, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 24)

cbind(beta = coef(m5))
exp(cbind(OR = coef(m5)))
```
### Overall effect of education on LBW

- Above results show that   
  - e1 & e3 difference significant
  - e1 & e2 difference is not significant
- What is the overall effect of maternal education on LBW?  

```{r}
#| echo: true
wald.test(b = coef(m5), Sigma = vcov(m5), Terms = 2:3)
```

### Is the difference between e2 & e3 significant?

- We already have the results for e1 vs e2 and e1 vs e3

```{r}
l <- cbind(0, 1, -1)
wald.test(b = coef(m5), Sigma = vcov(m5), L = l) |> print.wald.test()
```
