---
title: "Bias/Measurement Error I & II"
format: 
  clean-typst:
    include-in-header: "custom.typ"
    commands: [small-cite]
execute: 
  echo: false
  warning: false    
author:
  - name: A Pathmeswaran
    orcid: 0000-0003-4065-2639
    email: pathmes@kln.ac.lk
date: 2026-03-20
date-format: long
handout: false
---

```{r}
#| label: loadPacks

library(tinytable)
# library(tidyverse)
# library(aod)
# library(summarytools)
# library(gt)
# library(gtsummary)

```

### Syllabus

-   Validity
    -   Internal
    -   External
-   Bias
    -   Selection
    -   Information Bias

{{< v 2em >}}

-   What is the relationship between the terms validity & bias?

### Definitions

::: callout-note
## Bias

Deviation of results from the truth or the process leading to such deviation.
:::

{{< v 2em >}}

::: callout-note
## Validity

-   Measurement - the degree to which a measurement measures what it purports to measure.
-   Study - the degree to which the inferences drawn from a study are true.
:::

### Internal validity

Closeness of the results/ study findings to the truth about study participants.

-   Clinical trials – likely to have high internal validity
-   Observational analytic studies – careful assessment needed

. . .

{{< v 2em >}}

-   Descriptive studies – relevance of internal validity?

### Internal Validity

Internal validity is affected by

-   Selection bias
-   Information bias
-   Confounding

. . .

{{< v 2em >}}

-   Selection bias in cohort studies?

### External validity/ generalisability

{{< v 2em >}}

-   Applicability of the findings to the reference population (in a descriptive study)

-   Applicability of findings to other populations/ others (clinical trials & analytical observational studies)

### Clinical Trials

-   explanatory - Can the intervention work?

-   pragmatic - Does the intervention work in routine care?

-   Internal & external validity of explanatory and pragmatic trials

-   Can there be external validity without internal validity?

### Reliability

**Reliable** - free of error

-   Information obtained
-   Study findings

### Reliability & measurement error

All measurements are associated with measurement error (misclassification for categorical variables)

Maybe due to -

-   Poor quality measurement instrument
-   Subjects
-   Person taking measurements
-   Measurement process

### Importance / Usefulness of Tools

-   Screening instruments
    -   Objective measures of some construct
-   Numeric outcome measures in clinical trials

------------------------------------------------------------------------

-   Is it measuring something?

. . .

-   Consistent results -\> Reliability

-   Does it measure what it is supposed to measure?

    -   Validity

### Measurement Error

{{< v 2em >}}

$$\text {Measurement} =\text {Actual value} + \text {Measurement error}$$

{{< v 2em >}}

. . .  

- Define measurement

### Reliability

{{< v 2em >}}

$$\text {Reliability} =\frac{\text {Subject Variability}} {\text {Subject Variability}+ \text {Measurement error}}$$

ICC - Intraclass Correlation Coefficient

### Data for ICC

::::: {layout-ncol="2"}
::: {#first-column}
[Data]{.alert}

```{r}
O1 <- c(6, 4, 2, 3, 5, 8, 5, 6, 4, 7)
O2 <- c(7, 5, 2, 4, 4, 9, 7, 7, 6, 9)
O3 <- c(8, 6, 2, 5, 6, 10, 9, 8, 8, 8)

as.data.frame(cbind(O1, O2, O3)) |> tt() |>  tt_style(j = 1:3, align = 'r')
```
:::

::: {#second-column}
[Summary]{.alert}

```{r}
print(paste("O1 - Mean =", mean(O1), "; SD", round(sd(O1), 1)))
print(paste("O2 - Mean =", mean(O2), "; SD", round(sd(O2), 1)))
print(paste("O3 - Mean =", mean(O3), "; SD", round(sd(O3), 1)))
# print(paste("Mean O2 =", mean(O2)))
# print(paste("Mean O3 =", mean(O3)))
# 
# print(paste("SD O1 =", round(sd(O1), 1)))

```
:::
:::::

### ICC

```{r}
irr::icc(cbind(O1, O2, O3), "twoway")
```

### Binary variable

| X     | Yes |  No | Total |
|-------|----:|----:|------:|
| Yes   | 140 |  10 |   150 |
| No    |  10 | 140 |   150 |
| Total | 150 | 150 |   300 |

: Two observers

$$\text{Observed agreement} = \frac{140 + 140}{300} = 0.933$$

. . .  

$$ \text{Expected agreement} = \frac{\frac{150 \times 150}{300} + \frac{150 \times 150}{300}}{300} = 0.5$$


$$\kappa = \frac {P_o - P_e}{1.0 - P_e}$$


### Another Binary variable

| X     | Yes |  No | Total |
|-------|----:|----:|------:|
| Yes   | 270 |  10 |   280 |
| No    |  10 |  10 |    20 |
| Total | 280 |  20 |   300 |

- Observed agreement
- Expected agreement
- Kappa

### Cronbach's Alpha

- Based on item total correlation
- A measure of internal consistency
- Varies from 0 to 1
- Above 0.7 good value for surveys
- Instruments used in clinical practice should have values above 0.9
