---
title: "Logistic Regression III & IV"
format: 
  clean-typst:
    include-in-header: "custom.typ"
    commands: [small-cite]
execute: 
  echo: false
  warning: false    
author:
  - name: A Pathmeswaran
    orcid: 0000-0003-4065-2639
    email: pathmes@kln.ac.lk
date: 2026-03-16
date-format: long
handout: false
brand:
  typography: 
    fonts: 
      - family: Roboto
        weight: [100, 200, 300, 400, 500, 600, 700]
        style: [normal, italic]
        source: google
    base: Roboto
    headings: Roboto
---

```{r}
#| label: loadPacks
library(tidyverse)
library(summarytools)
library(gt)
library(gtsummary)
library(aod)
library(texreg)
library(rcompanion)
# library(generalhoslem)
```


### Syllabus
- Logistic regression model $\checkmark$
- interpretation of coefficients for dichotomous & continuous variables $\checkmark$
- Multiple logistic regression:  
  a. Wald test $\checkmark$
  b. Likelihood ratio test
  c. Dummy variables $\checkmark$
  d. Model building strategies
  e. Goodness of fit tests
  


### Log likelihood ratio test
- Log likelihood is a negative number
- A model with a better fit will have a log likelyhood closer to zero
$$LR = -2ln(\frac{L(m_1)}{L(m_2)})= 2(loglik(m_2) - loglik(m_1))$$


### Definition - AIC

::: callout-note
## AIC - Akaike information criterion 

$$\displaystyle \mathrm {AIC} \,=\,2k-2\ln({\hat {L}})$$
k is the number of parameters in the model. 

:::
Lower values of $\displaystyle \mathrm -2\ln({\hat {L}})$ indicate better fit. As we add more parameters to the model the $\displaystyle \mathrm -2\ln({\hat {L}})$ tends to decrease. In the calculation of AIC, the addition of 2k penalises the inclusion of additional parameters to the model. 


### Definition - BIC

::: callout-note
## BIC - Bayesian information criterion 

$$\displaystyle \mathrm {BIC} =k\ln(n)-2\ln({\widehat {L}})$$
k is the number of parameters in the model. 
n is the sample size
:::
It can be seen that the penalty for additional parameters is higher for BIC than for AIC. 


### Log likelihood test (predictor - education)

```{r}
load("C:/AP/MyR/lectures/PGIM2/MD_CM_2023/md2023stats/bwt.RData")
bwt$lbw <- bwt$bw < 2500
bwt$ht_cm <- bwt$height*100
bwt$edu <- as.character(bwt$education)
bwt$edu3 <- case_match(
bwt$edu,
c("TEC", "GDU", "PG", "1-5", "6-8") ~ "e1",
c("8-10", "OL") ~ "e2",
c("AL", "NONE" ) ~ "e3"
)

bwt$edu3[bwt$idn == 54] <- "e3"
bwt$edu3[bwt$idn == 66] <- "e3"
bwt$edu3[bwt$idn == 182] <- "e3"
bwt$edu3[bwt$idn == 239] <- "e3"
bwt$edu3[bwt$idn == 259] <- "e3"
bwt$edu3[bwt$idn == 261] <- "e3"
bwt$edu3[bwt$idn == 432] <- "e3"


```


#### Likelihood ratio test
```{r}
m5 <- glm(
lbw ~ edu3,
data = bwt,
family = "binomial")

logLik(m5)

with(m5, pchisq(null.deviance - deviance, df.null - df.residual, lower.tail = FALSE))
```
### Pseudo R squared

```{r}
nagelkerke(m5)$Pseudo.R.squared.for.model.vs.null
```

### Maternal height and LBW

```{r}
m2 <- glm(
  lbw ~ ht_cm, 
  data = bwt, 
  family = "binomial")

gtsummary::tbl_regression(m2, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 24)


```



### Log likelihood test (predictor - height)


```{r}

logLik(m2)
with(m2, pchisq(null.deviance - deviance, df.null - df.residual, lower.tail = FALSE))

```
### Pseudo R squared

```{r}
nagelkerke(m2)$Pseudo.R.squared.for.model.vs.null
```
### Comparison of edu and height as predictors

**maternal edu**

```{r}
broom::glance(m5) |> tinytable::tt()
```

**maternal height**

```{r}
broom::glance(m2) |> tinytable::tt()
```


### LBW by Education & Maternal Height - Logistic Regression

```{r}
m6 <- glm(
lbw ~ edu3 + ht_cm,
data = bwt,
family = "binomial")

gtsummary::tbl_regression(m6, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 24)

```
### LBW by Education & Maternal Height - Logistic Regression - Coefficients

```{r}
cbind(beta = coef(m6), confint(m6))
exp(cbind(OR = coef(m6), confint(m6)))
```

## Log likelihood test (predictors(height & education))

### Likelihood ratio test
```{r}

logLik(m6)
with(m6, pchisq(null.deviance - deviance, df.null - df.residual, lower.tail = FALSE))

nagelkerke(m6)$Pseudo.R.squared.for.model.vs.null
```



### Other predictors

```{r}
bwt |> 
  select(weight, midacir, fundht) |> 
  tbl_summary(
     type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{median} ({p25}, {p75})", "{mean}, ({sd})")
  ) |> 
  as_gt() |> 
tab_options(table.font.size = 24)
```



### Comparison of models

```{r}
m4 <- glm(
lbw ~ weight,
data = bwt,
family = "binomial")

m7 <- glm(
lbw ~ midacir,
data = bwt,
family = "binomial")

m8 <- glm(
lbw ~ fundht,
data = bwt,
family = "binomial")

```




### Comparison (Fundal height +)

```{r}
f1 <- glm(
lbw ~ fundht + ht_cm,
data = bwt,
family = "binomial")

f2 <- glm(
lbw ~ fundht + weight,
data = bwt,
family = "binomial")

f3 <- glm(
lbw ~ fundht + midacir,
data = bwt,
family = "binomial")
```

```{r}
#| results = "asis"
texreg(list(m8, f1, f2, f3))
```

### Pseudo R squared for a few models {.scrollable}

```{r}
nagelkerke(m8)$Models[1]
nagelkerke(m8)$Pseudo.R.squared.for.model.vs.null

nagelkerke(f1)$Models[1]
nagelkerke(f1)$Pseudo.R.squared.for.model.vs.null

nagelkerke(f2)$Models[1]
nagelkerke(f2)$Pseudo.R.squared.for.model.vs.null

nagelkerke(f3)$Models[1]
nagelkerke(f3)$Pseudo.R.squared.for.model.vs.null
```

### Goodness of fit

- p > 0.05 No evidence of poor fit
```{r}


m2$formula
generalhoslem::logitgof(bwt$lbw, fitted(m2))

m8$formula
generalhoslem::logitgof(bwt$lbw, fitted(m8), g = 3)

f2$formula
generalhoslem::logitgof(bwt$lbw, fitted(f2))

```




## Is the improvement significant?
### Should we add weight as a predictor?
```{r}

anova(m8, f2, test='LR')

```

## Is the improvement significant?
### Should we add mid arm circumference as a predictor?
```{r}

anova(m8, f3, test='LR')

```

### LBW by Education & Maternal Height interaction - Logistic Regression

```{r}
m11 <- glm(
lbw ~ edu3 * ht_cm,
data = bwt,
family = "binomial")

gtsummary::tbl_regression(m11, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 24)

```


### Modelling strategy
- Depends on the outcome variable - 
  - binary - logistic regression
  - continuous - linear regression
  - time to event - Cox model
  
- Decide on the predictors
  - study hypothesis
  - adjusting for possible confounding
  - selection of variables is based on prior knowledge and the results of bivariate analysis
  - avoid automatic selection methods
  - recoding or tranforming of the predictors
- look at model statistics
  



### Boys smoking and cough 
```{r}
source(here::here("premdstat", "bland_boys_cough.R"))

    tbl_cross(t17l,
              row = boys,
              col = cough,
              percent = c("row")) |> 
  add_p() |> 
  as_gt() |> 
tab_options(table.font.size = 24)

```

### Parents smoking and cough among boys

```{r}
    tbl_cross(t17l,
              row = parents,
              col = cough,
              percent = c("row")) |> 
  add_p() |> 
  as_gt() |> 
tab_options(table.font.size = 24)
```
### Parents smoking and boys smoking

```{r}
    tbl_cross(t17l,
              row = boys,
              col = parents,
              percent = c("column")) |> 
  add_p() |> 
  as_gt() |> 
tab_options(table.font.size = 24)
```
### Cough vs boys smoking


```{r}
mbs <- glm(
cough ~ boys,
data = t17l,
family = "binomial")

broom::tidy(mbs) |> tinytable::tt() |> tinytable::format_tt(j = 2:4, digits = 3)

wald.test(b = coef(mbs), Sigma = vcov(mbs), Terms = 2:4)
```

### Cough vs boys smoking - Model stats

```{r}
broom::glance(mbs)[c(1, 6, 2, 7)] |> tinytable::tt()
broom::glance(mbs)[c(3, 4, 5, 8)]|> tinytable::tt()
```

### Cough vs parents smoking


```{r}
mps <- glm(
cough ~ parents,
data = t17l,
family = "binomial")

broom::tidy(mps) |> tinytable::tt() |> tinytable::format_tt(j = 2:4, digits = 3)

wald.test(b = coef(mps), Sigma = vcov(mps), Terms = 2:3)
```

### Cough vs parents smoking - Model stats

```{r}
broom::glance(mps)[c(1, 6, 2, 7)] |> tinytable::tt()
broom::glance(mps)[c(3, 4, 5, 8)]|> tinytable::tt()
```

### Cough vs boys & parents smoking


```{r}
ms2 <- glm(
cough ~ boys + parents,
data = t17l,
family = "binomial")

broom::tidy(ms2) |> tinytable::tt() |> tinytable::format_tt(j = 2:4, digits = 3)
wald.test(b = coef(ms2), Sigma = vcov(ms2), Terms = 2:4)
wald.test(b = coef(ms2), Sigma = vcov(ms2), Terms = 5:6)

```

### Cough vs boys & parents smoking - Model stats

```{r}
broom::glance(ms2)[c(1, 6, 2, 7)] |> tinytable::tt()
broom::glance(ms2)[c(3, 4, 5, 8)]|> tinytable::tt()
```

### Cough vs boys & parents smoking (interaction)


```{r}
ms2i <- glm(
cough ~ boys * parents,
data = t17l,
family = "binomial")

broom::tidy(ms2i) |> tinytable::tt() |> tinytable::format_tt(j = 2:4, digits = 3)
wald.test(b = coef(ms2i), Sigma = vcov(ms2i), Terms = 2:4)
wald.test(b = coef(ms2i), Sigma = vcov(ms2i), Terms = 5:6)
wald.test(b = coef(ms2i), Sigma = vcov(ms2i), Terms = 7:12)

```

### Cough vs boys & parents smoking i - Model stats

```{r}
broom::glance(ms2i)[c(1, 6, 2, 7)] |> tinytable::tt()
broom::glance(ms2i)[c(3, 4, 5, 8)]|> tinytable::tt()
```

