---
title: "Analysis of Cohort Studies I & II	"
format: 
  clean-typst:
    include-in-header: "custom.typ"
    commands: [small-cite]
execute: 
  echo: false
  warning: false    
author:
  - name: A Pathmeswaran
    orcid: 0000-0003-4065-2639
    email: pathmes@kln.ac.lk
date: 2026-04-02
date-format: long
bibliography: [static/references.bib]
bibliographystyle: "apa"
handout: false
---

```{r}
#| label: loadPacks

# library(ggdag)
library(tinytable)
# library(tidyverse)
# library(aod)
# library(summarytools)
# library(gt)
# library(gtsummary)
# suppressPackageStartupMessages(library(kableExtra))


```

### Syllabus

-   Relative Risk (RR) Attributable Risk (AR)\
-   Incidence Rate Ratio (IRR) Incidence Rate Difference (IRD)
-   Attributable Risk % (AR%) Population Attributable Risk (PAR) Population Attributable Risk Percent (PAR%)
-   Preventable fraction\
-   Calculation of 95% CIs
    -   Mantel-Haenszel method
    -   Test-based method
-   Calculation of AR & PAR% in case-control studies

### Closed vs Open Cohorts

**Calculation of incidence rate**

-   Numerator = Number of cases during the follow-up period
-   Denominator
    -   Closed cohort - Number at the start of follow-up
    -   Open cohort - person time

### Q1

```{r}
source(here::here("dmacm.R"))
tinytable::tt(dmacm, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")
```

The data is from a cohort study [@kasturiratne2020]. It shows the association between baseline diabetes status and mortality over 10 years of follow-up.

-   Calculate the following
    -   Relative risk
    -   Attributable risk

### Relative Risk

$$\text{Relative Risk} = \frac{\text{Frequency among the exposed}}{\text{Frequency among the unexposed}}$$

-   Frequency

    -   Incidence or Mortality

-   Range for RR: 0 to inf

-   RR = 1 No association

-   RR \> 1 Exposure is positively related to outcome.

-   RR \< 1 Exposure protects against outcome (Vaccination and disease)

### Risk Difference

-   Attributable risk – Rate of disease in the exposed that can be attributed to the exposure

$$\text{Attributable Risk}_{(exp)} = 
            \text{Incidence}_{(exp)} - \text{Incidence}_{(unexp)}$$

-   Population excess rate - Rate of disease in the population that can be attributed to the exposure

$$\text{Population excess rate} =
            \text{Incidence}_{(pop)} - \text{Incidence}_{(unexp)}$$

### Relative Risk vs Attributable Risk

-   Relative Risk
    -   Strength of association
    -   Aetiology

{{< v 2em >}}

-   Population excess rate
    -   Used in planning/management
    -   Estimate absolute numbers

{{< v 2em >}}

-   Attributable Risk ($\text{Attributable Risk}_{(exp)}$)

### Assumptions underlying statistical methods

- Independence of outcome
  - Within a group, Knowing the outcome status of one group member does not provide any information regarding the outcome status of another group member.
- Homogeneity of risk
  - Within a group, all person time units have the same risk or probability of outcome. $\text{2 persons}\times{\text{50 years}} = \text{20 persons}\times{\text{5 years}}$
  - Members of a group have identical risk.
  

### Calculation of 95% CI
#### Person-Time  data 

```{r}
tinytable::tt(t141, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")

```
- Incidence Rate - $I1 = A1/T1$ ; $I0 = A0/T0$  

- Incidence Ratio - $IR  = I1/I0 = \frac{A1/T1}{A0/T0}$  

- Incidence Difference - $ID = I1 - I0 ={A1/T1} - {A0/T0}$


### Calculation of 95% CI
#### Person-Time  data 

```{r}
tinytable::tt(t141, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")

```

- Incidence Ratio - $IR  = I1/I0 = \frac{A1/T1}{A0/T0}$  

When the null hypothesis is true, I1 = I0.    
- E the average number of exposed cases is $E = M\times T1/T$

- V the variance of A1 is $V = \frac{E \times T0}{T} = \frac{M \times T1 \times T0}{T^2}$  


$$X~score = \frac{A1 - E}{\sqrt V}$$

### Calculation of 95% CI
#### Person-Time  data 

```{r}
tinytable::tt(t141, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")

```

$$SE[ln(IR)] = \sqrt{(\frac{1}{A1} + \frac{1}{A0})}$$

$$SE(ID) = \sqrt{(\frac{A1}{T1^2} + \frac{A0}{T0^2})}$$

### Breast cancer following multiple exposure to radiation among patients with TB (Table 14-2 Rothman)

{{< v 2em >}}

```{r}
tinytable::tt(t142, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white") |> 
  tinytable::format_tt(j= 2:4, digits = 3, num_mark_big =  " ")

```

Calculate -   

1. IR
2. ID
3. $X~score$
4. SE[ln(IR)] and the 95% CI for IR
5. SE(ID) and the 95% CI for ID

### Breast cancer following multiple exposure to radiation among patients with TB (Table 14-2) - Answers 1, 2, 3

{{< v 2em >}}

```{r}
tinytable::tt(t142, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white") |> 
  tinytable::format_tt(j= 2:4, digits = 3, num_mark_big =  " ")

```

Answers -   

1. $IR = \frac{41/28,010}{15/19,017} = 1.86$
2. ID = 67.5 per 100 000 years
3. X~score 
  - $E = \frac{56(28,010)}{47, 027} = 33.35$
  - $V = \frac{33.35(19,017)}{47, 027} = 13.49$
  - $X~score = \frac{41 - 33.35}{\sqrt 13.49}$

### Breast cancer following multiple exposure to radiation among patients with TB (Table 14-2) - Answers 4, 5

{{< v 2em >}}

```{r}
tinytable::tt(t142, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white") |> 
  tinytable::format_tt(j= 2:4, digits = 3, num_mark_big =  " ")

```

Answers 

- $SE[ln(IR)] =  \sqrt{(\frac{1}{41} + \frac{1}{15})} = 0.302$
- 95% CI for ln(IR) = $ln(1.86) \pm 1.96(0.302)$
- 95% CI for IR = $exp[ln(1.86) \pm 1.96(0.302)] = \text{1.03 to 3.35}$

{{< v 1em >}}

- $SE(ID) = \sqrt{(\frac{41}{28,010^2} + \frac{15}{19,017^2})} = $ 30.6 per 100,000 years
- 95% CI for ID = $67.5 \pm 1.96(30.6) = \text{7.5 to 127.5}$ per 100,000 years




### Calculation of 95% CI
#### Count  data 

```{r}

tinytable::tt(t143, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")

```
- Risk - $R1 = A1/N1$ ; $R0 = A0/N0$  

- Risk Ratio - $RR  = R1/R0 = \frac{A1/N1}{A0/N0}$  

- Risk Difference - $RD = R1 - R0 ={A1/N1} - {A0/N0}$


### Calculation of 95% CI
#### Count  data 

```{r}

tinytable::tt(t143, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")

```
- Risk Ratio - $RR  = R1/R0 = \frac{A1/N1}{A0/N0}$  

- If R1 = R0, the average number of exposed cases (E) when the total cases is M1 - E = M1N1/N
- Variance of A = V
  - $V = \frac{EM0N0}{N(N-1)} =  \frac{M1N1M0N0}{N^2(N-1)}$
  
- $X~score = \frac{A1 - E}{\sqrt V}$
 
### Calculation of 95% CI
#### Count  data 

```{r}

tinytable::tt(t143, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")

```

$$SE[ln(RR)] = \sqrt{(\frac{1}{A1} - \frac{1}{N1} +
\frac{1}{A0} - \frac{1}{N0} )}$$

$$SE(RD) = \sqrt{(\frac{A1B1}{N1^2(N1 - 1)} + \frac{A0B0}{N0^2(N0 - 1)})}$$

### Antibodies in breast milk and diarrea in infants (14-4)

```{r}
tinytable::tt(t144, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")
```

Calculate -   

1. RR
2. RD
3. $X~score$
4. SE[ln(RR)] and the 95% CI for RR
5. SE(RD) and the 95% CI for RD


### Diarrhea in infants 

```{r}

tinytable::tt(t144, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")

```
- Risk Ratio - $RR  = R1/R0 = \frac{A1/N1}{A0/N0}$  

-  E = M1N1/N
- $V = \frac{EM0N0}{N(N-1)} =  \frac{M1N1M0N0}{N^2(N-1)}$
  
- $X~score = \frac{A1 - E}{\sqrt V}$
 
### Diarrhea in infants 


```{r}

tinytable::tt(t144, rownames = TRUE) |> 
  tinytable::style_tt(j = 2:4, align = "r") |> 
  tinytable::style_tt(j = 1, i = 0, color = "white")

```

$$SE[ln(RR)] = \sqrt{(\frac{1}{A1} - \frac{1}{N1} +
\frac{1}{A0} - \frac{1}{N0} )}$$

$$SE(RD) = \sqrt{(\frac{A1B1}{N1^2(N1 - 1)} + \frac{A0B0}{N0^2(N0 - 1)})}$$


### Attributable fraction in case control studies

Exposure Attributable Fraction (EAF)  = $\frac{OR - 1}{OR}$

Population Attributable Fraction (PAF) = $\frac{P \times (OR - 1)}{P \times (OR - 1) + 1}$

P - prevalence of exposure in the base population

### Questions

1. Is SMR (standardised mortality ratio) a rate ratio?
2. Are these two formulae the same  
$$SE(RD) = \sqrt{(\frac{A1B1}{N1^2(N1 - 1)} + \frac{A0B0}{N0^2(N0 - 1)})}$$  

$$SE_{diff} = \sqrt{\frac{p_1q_1}{n_1}+\frac{p_2q_2}{n_2}}$$



### References
