---
title: "Logistic Regression I & II"
author: "A Pathmeswaran<br>pathmes@kln.ac.lk<br>2025-02-28"
format: 
  revealjs:
   theme: [default, pgim2025.scss]
   footer: "MD Community Medicine 2023"
   slide-number: c/t
   embed-resources: true
   standalone: true
---

```{r}
#| label: loadPacks

library(summarytools)
library(gt)
library(gtsummary)

```

```{r}
#| label: getData

source(here::here("kirkwood19.2.R"))
```


## Syllabus
- Logistic regression model
- interpretation of coefficients for dichotomous & continuous variables

 . . .  
 
- Multiple logistic regression:  
  a. Wald test
  b. Likelihood ratio test
  c. Dummy variables
  d. Model building strategies
  e. Goodness of fit tests
  
## Logit

Logit is equal to log of odds  
If p is a proportion  

$$logit(p) = log_e (\frac{p}{1-p})$$

## Data from Kirkwoods book

```{r}
set.seed(21022025)
t3[sample(nrow(t3), 7), ] |> 
  kableExtra::kable(row.names = FALSE)
```

## MF postivity by area

```{r}
ctable(x = t3$Area,
         y = t3$Microfilaria,
       totals = F)
```
```{r}
tx <- table(t3$Area, t3$Microfilaria)
osv <- tx[1,2]/tx[1,1]
orf <- tx[2,2]/tx[2,1]
```

 . . .  
 
Odds of MF positivity in Rainforest - `r tx[2,2]` / `r tx[2,1]` = 
`r round(orf, 2)`  
Odds of MF positivity in Savannah - `r tx[1,2]` / `r tx[1,1]` = 
`r round(osv, 2)`


## Table with OR

```{r}
or_fun <- function(data, variable, by, ...) {
  table(data[[by]], data[[variable]]) %>%
    fisher.test() %>%
    broom::tidy()
}

  t3 %>%
  select(Microfilaria, Area) %>%
  tbl_summary(by = Area, missing = "no", type = all_dichotomous() ~ "categorical") %>%
  add_difference(test = Microfilaria ~ or_fun, estimate_fun = Microfilaria ~ style_ratio) %>%
  modify_header(estimate ~ "**Odds Ratio**") |> 
  as_gt() |> 
tab_options(table.font.size = 32)
```

## Properties / Advantages of Odds Ratios

1. When the outcome is rare the odds ratio is the same as the risk ratio
2. The conclusions are identical whether the outcome is the occurence of an event or the absence of the event.
2. Unlike the risk ratio the odds ratio is not constrained when the outcome is common in the unexposed group.

## MF postivity by area 2

```{r}
ctable(x = t3$Area,
         y = t3$Microfilaria,
         OR = TRUE,
         chisq = TRUE)
```

## Logistic regression{.smaller}

### Odds Ratio in Case Control Studies
$$\text{Odds Ratio} = \frac{\text{Odds of Exposure among Cases}}
{\text{Odds of Exposure among Controls}}$$
Same as  
$$\text{Exposure Odds Ratio} = \frac{\text{Odds in Exposed Group}}
{\text{Odds in Unexposed Group}}$$

$$\text{Odds in Exposed Group} = \text{Odds in Unexposed Group} \times \text{Exposure Odds Ratio}$$

## Logistic regression output


::: {.panel-tabset}

### Raw

```{r}
m1 <- glm(
  Microfilaria ~ Area, 
  data = t3, 
  family = "binomial")
summary(m1)
```

### OR

```{r}
exp(cbind(OR = coef(m1), confint(m1)))
```

### Print 1

```{r}
gtsummary::tbl_regression(m1)|> 
  as_gt() |> 
tab_options(table.font.size = 32)
```


### Print 2

```{r}
gtsummary::tbl_regression(m1, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 32)
```

:::



## Birth Weight Data

```{r}
load("C:/AP/MyR/lectures/PGIM2/MD_CM_2023/md2023stats/bwt.RData")
bwt$lbw <- bwt$bw < 2500
bwt$ht_cm <- bwt$height*100
bwt$ht_dm <- bwt$height*10

bwt |> select(height, bw, lbw) |> 
  tbl_summary(
     type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{median} ({p25}, {p75})", "{mean}, ({sd})")
  ) |> 
  as_gt() |> 
tab_options(table.font.size = 32)
```

## Maternal height and LBW

```{r}
m2 <- glm(
  lbw ~ ht_cm, 
  data = bwt, 
  family = "binomial")

gtsummary::tbl_regression(m2, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 32)

cbind(beta = coef(m2), confint(m2))
exp(cbind(OR = coef(m2), confint(m2)))
```

## Maternal height (in dm) and LBW

```{r}
m3 <- glm(
  lbw ~ ht_dm, 
  data = bwt, 
  family = "binomial")

gtsummary::tbl_regression(m3, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 32)

exp(cbind(OR = coef(m3), confint(m3)))
```

## LBW Predicted Probability 

$$p = \frac{odds}{(1 + odds)}$$
$$\text{Odds in Exposed Group} = \text{Odds in Unexposed Group} \times \text{Exposure Odds Ratio}$$

Calculate the probability of LBW for mother's height = 140

## LBW Predicted Probability for selected maternal heights

```{r}
h <- seq(130, 190, 10)
cbind(h, exp(predict(m2, data.frame(ht_cm = h))))


```

## What if we had used a linear model
```{r}
mg <- glm(
lbw ~ ht_cm,
data = bwt,
family = "gaussian")

gtsummary::tbl_regression(mg, exponentiate = TRUE)|> 
  as_gt() |> 
tab_options(table.font.size = 32)

cbind(beta = coef(mg), confint(mg))
exp(cbind(OR = coef(mg), confint(mg)))

```

## LBW Predicted Probability (based on linear regression) for selected maternal heights

 . . .  
 
 
```{r}
cbind(h, exp(predict(mg, data.frame(ht_cm = h))))
```
