---
title: "Statistics 1 & 2"
author: "A Pathmeswaran<br>pathmes@kln.ac.lk<br>2023-01-31"
format: 
  revealjs:
   theme: [default, custom.scss]
   footer: "MD Community Medicine 2023"
   embed-resources: true
   standalone: true
---

```{r}
#| label: load-packages
suppressMessages(library(tidyverse))
suppressMessages(library(readr))
suppressMessages(library(broom))
suppressMessages(library(flextable))
suppressMessages(library(huxtable))
```

## Aim

-   To enable students to comprehend, apply and interpret the results of a range of advanced techniques for the design and analysis of epidemiological studies.

## Objectives

-   Be able to:

    1.  describe the basis of relevant statistical tests with underlying assumptions in the analysis of epidemiological data\
    2.  develop skill of applying appropriate statistical and computational methods in the current scientific applications\
    3.  interpret and critique the statistical applications used in relevant literature

## Outline

-   Parametric tests:
    a.  T-test: paired & independent t-test
    b.  Chi-squared test
    c.  Z-test for means and proportions
    d.  Estimation of population parameters
    e.  Assessing normal distribution

## Q1

A researcher claims to have developed a herbal product to prevent complications of dengue. The use of this product early in the illness for five days had resulted in a 10% increase in the platelet count.

1)  List all relevant information needed to assess the validity of this claim.

## Q1 - Summary data

The above claim was based on an observational study of 30 patients with dengue fever (antigen test positive). The platelet values on day 0 and day 5 are summarised below -

```{r}
#| label: q1-data
#| echo: false
#| warning: false
#| message: false

q0l <- read_csv(here::here("data", "q0l.csv"))
t1 <- q0l %>%
  group_by(Day) %>%
  summarise(mean = mean(Platelets), sd = sd(Platelets))

huxtable(t1) %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_number_format(row = everywhere, col = 3, value = "%5.1f" ) %>%   
set_width(0.8) %>%
set_caption("Platelets on Day 0 and Day 5 - Summary ")

```

## Q1 analysis 1

```{r}
#| label: q1-analysis1
#| echo: false
#| warning: false
#| message: false

q0data <- read_csv(here::here("data", "q0data.csv"))
t1 <- tidy(t.test(q0data$plt_d5, q0data$plt_d0, paired = FALSE, var.equal = TRUE)) 

t1[ , 1:6] %>%   
huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8) 
 
t1[ , 7:9] %>% 
 huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8)

```

**Interpretation?**

## Q1 analysis 2

```{r}
#| label: q1-analysis2
#| echo: false
#| warning: false
#| message: false

t1 <- tidy(t.test(q0data$plt_d5, q0data$plt_d0, paired = FALSE, var.equal = FALSE))

t1[ , 1:6] %>%   
huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8) 
 
t1[ , 7:9] %>% 
 huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8)

```

## How many types of t tests are there?

-   Describe the different types of t tests

## Q1 - Summary data 2

```{r}
#| label: q1-data2
#| echo: false
#| warning: false
#| message: false

q0l <- read_csv(here::here("data", "q0l.csv"))
t1 <- q0l %>%
  group_by(Day) %>%
  summarise(mean = mean(Platelets), sd = sd(Platelets))

huxtable(t1) %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_number_format(row = everywhere, col = 3, value = "%5.1f" ) %>%  
set_width(0.8) 

q0w <- read_csv(here::here("data", "q0data.csv"))
t12 <- q0w %>% 
  summarise(Diff_m = mean(diff), Diff_sd = sd(diff))

huxtable(t12) %>%
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_number_format(row = 2, col = everywhere, value = "%5.1f" ) %>%
set_width(0.8)
```

## Results of one sample t tests

-   Mean difference = 0 (Null hypothesis)

```{r}
#| label: q1-analysis3
#| echo: false
#| warning: false
#| message: false

t2 <- tidy(t.test(q0data$diff, mu = 0)) 
  t2[ , 1:4] %>%   
huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8) 
 
t2[ , 5:7] %>%
 huxtable() %>%
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8)

```

## Results of paired t test

```{r}
#| label: q1-analysis4
#| echo: false
#| warning: false
#| message: false

q0l <- read_csv(here::here("data", "q0l.csv"))
t4 <- tidy(t.test(q0l$Platelets~q0l$Day, paired = TRUE)) 

t4[ , 1:4] %>%   
huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8) 
 
t4[ , 5:7] %>%
 huxtable() %>%
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8)
```

## Results of paired t test

```{r}
#| label: q1-analysis4r
#| echo: false
#| warning: false
#| message: false

q0data <- read_csv(here::here("data", "q0data.csv"))
t4 <- tidy(t.test(Pair(plt_d5, plt_d0) ~ 1, data = q0data)) 

  t4[ , 1:4] %>%   
huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8) 
 
t4[ , 5:7] %>%
 huxtable() %>%
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8)
```

## Scientific Notation {.larger}

**Scientific Notation**

P value = `r format(t4$p.value, scientific = TRUE)`

**Decimal Form**

P value = `r format(t4$p.value, scientific = FALSE)`

P value = `r format(t4$p.value, scientific = FALSE, digits = 1)`

**SPSS Output**  

P value = 0.000

**Sensible Presentation**  

P < 0.001

## Data from a clinical trial

```{r }
#| label: yop-data
#| echo: FALSE
 
yop <- tribble(
~status, ~charcoal, ~ placebo,
"Dead", 5,           16,
"Alive", 196,        184,
"Total", 201,        200
)
yop %>% 
kableExtra::kbl()

p1 <- yop[1,2]/yop[3,2]
q1 <- 1-p1
p2 <- yop[1,3]/yop[3,3]
q2 <- 1-p2
p <- (yop[1,2]+yop[1,3])/(yop[3,2]+yop[3,3])
q <- 1-p


```

de Silva, H.A., Fonseka, M.M.D., Pathmeswaran, A., Alahakone, D.G.S., Ratnatilake, G.A., Gunatilake, S.B., Ranasinha, C.D., Lalloo, D.G., Aronson, J.K., de Silva, H.J., 2003.\
Multiple-dose activated charcoal for treatment of yellow oleander poisoning: a single-blind, randomised, placebo-controlled trial. Lancet 361, 1935--1938.

## Chi Squared Test

```{r }
#| echo: false
#| label: chi-sqr
tidy(chisq.test(yop[1:2, 2:3])) %>% 
  mutate_if(is.numeric, format, digits = 3) %>% 
 kableExtra::kbl() %>% 
  kableExtra::kable_styling(latex_options = c("scale_down"))  

```

-   Conditions for the Chi squared test?\
-   Is Chi squared a parametric test?

## Observed and Expected Values

```{r }
#| label: chi-sq-expct
#| echo: false
  
yop %>% 
kableExtra::kbl()

chisq.test(yop[1:2, 2:3])$expected %>% 
   kableExtra::kbl(digits = 1)  

```

## Comparing Two Proportions

```{r }
#| label: two-prop
#| echo: false
x1 <- c(yop$placebo[1], yop$charcoal[1])
n1 <- c(yop$placebo[3], yop$charcoal[3])

t5 <- tidy(prop.test(x1, n1))

t5[ , 1:4] %>%   
huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8) 

```

Difference: `r round((t5$estimate1 - t5$estimate2), 3)`\
95% CI: `r round(t5$conf.low, 3)` - `r round(t5$conf.high, 3)`
  
  
## Books  

Bland, M. (2015). **An introduction to medical statistics** (Fourth edition). Oxford University Press.  

Armitage, P., Berry, G., & Matthews, J. N. S. (2001). **Statistical methods in medical research** (4th ed). Blackwell Science.  


## Authors

Altman, Douglas G    
Bland, Martin  
Colins, Garry S  
Norman, Geoffrey R  
Riley, Richard D  
Senn, Stephen  
Streiner, David L  

