---
title: "Statistics 1 & 2"
author: "Prof A Pathmeswaran<br>pathmes@kln.ac.lk<br>2023-01-31"
format: 
  revealjs:
   theme: [default, custom.scss]
   footer: "MD Community Medicine 2023"

---

```{r}
#| label: load-packages
suppressMessages(library(tidyverse))
suppressMessages(library(readr))
suppressMessages(library(broom))
suppressMessages(library(flextable))
suppressMessages(library(huxtable))
```

## Aim

- To enable students to comprehend, apply and interpret the results of a range of advanced techniques
for the design and analysis of epidemiological studies.  

## Objectives  

- Be able to:  

    1. describe the basis of relevant statistical tests with underlying assumptions in the analysis of epidemiological data  
    2. develop skill of applying appropriate statistical and computational methods in the current scientific applications  
    3. interpret and critique the statistical applications used in relevant literature  

## Outline

- Parametric tests:
    a. T-test: paired & independent t-test
    b. Chi-squared test
    c. Z-test for means and proportions
    d. Estimation of population parameters
    e. Assessing normal distribution

## Q1
A researcher claims to have developed a herbal product to prevent complications of dengue.  The use of this product early in the illness for five days had resulted in a 10% increase in the platelet count.  

1) List all relevant information needed to assess the validity of this claim.

## Q1 - Summary data
The above claim was based on an observational study of 30 patients with dengue fever (antigen test positive).  The platelet values on day 0 and day 5 are summarised below - 

set_number_format(ht, row, col, value )
```{r}
#| label: q1-data
#| echo: false
#| warning: false
#| message: false

q0l <- read_csv(here::here("data", "q0l.csv"))
t1 <- q0l %>%
  group_by(Day) %>%
  summarise(mean = mean(Platelets), sd = sd(Platelets))

huxtable(t1) %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_number_format(row = everywhere, col = 3, value = "%5.1f" ) %>%   
set_width(0.8) %>%
set_caption("Platelets on Day 0 and Day 5 - Summary ")

```


## Q1 analysis 1

```{r}
#| label: q1-analysis1
#| echo: false
#| warning: false
#| message: false

q0data <- read_csv(here::here("data", "q0data.csv"))
t1 <- tidy(t.test(q0data$plt_d5, q0data$plt_d0, paired = FALSE, var.equal = TRUE)) 

t1[ , 1:6] %>%   
huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8) 
 
t1[ , 7:9] %>% 
 huxtable() %>% 
  set_bold(row = 1, col = everywhere) %>%
set_bottom_border(row = 1, col = everywhere) %>%
set_width(0.8)

```
