---
title: "ANOVA 5 & 6"
author: "A Pathmeswaran<br>pathmes@kln.ac.lk<br>2024-02-16"
format: 
  revealjs:
   margin: 0.1
   theme: [default, custom.scss]
   footer: "MD Community Medicine 2024"
   embed-resources: true
   standalone: true
---

## Syllabus

-   Factorial ANOVA
    -   Main effects & interaction
    -   Calculation
    -   Interpretation of two way ANOVA table
-   Completely randomized design
-   Randomised block design
-   Balanced incomplete block design,
-   **Latin square design**,
-   Repeated measure ANOVA, MANOVA, \***ANCOVA**

```{r}
#|label: load_packages
#|echo: false
#|message: false
#|warning: false

suppressMessages(library(tidyverse))
suppressMessages(library(gtsummary))
suppressMessages(library(broom))
suppressMessages(library(car))
suppressMessages(library(MASS))
suppressMessages(library(tibble))
suppressMessages(library(patchwork))
suppressMessages(library(cowplot))
suppressMessages(library(gt))
```

## ANCOVA Analysis of Covariance

-   Used in analysing data from baseline and followup measurements\
-   Three alternative methods of analysis of such data in a clinical trial
    -   Compare the groups using the final measurement as the dependent variable
    -   Compare the groups using the change in values as the dependent variable
    -   ANCOVA

## Trial on Voice therapy for vocalfold nodules

-   Outcome measure - voice quality assessed by a therapist on a scale from 0 to 120 - **Lower score better quality**
-   Intervention group 32 participants
-   Control group 32 participants

## Summary of baseline & followup scores

```{r}

dfw <- readRDS(here::here("data", "dfw.rds"))

dfw %>% dplyr::select(Group, Pre, Post) %>% 
  tbl_summary(by = Group,
              statistic = all_continuous() ~ "{mean} ({sd})", 
              digits = list(c(Pre, Post)~c(1, 1))) %>%
  as_gt() %>%
  tab_options(table.font.size = 36)

```

## Boxplot of baseline & followup scores

```{r}
p0 <- dfw %>% 
  pivot_longer(!c(serialno,Group), names_to = "Timing", values_to = "score") %>% 
  ggplot(aes(score, Timing, color = Group))
p00 <- p0 + geom_boxplot() 

p00

```

## Another Boxplot of baseline & followup scores

```{r}
p1 <- dfw %>% 
  pivot_longer(!c(serialno,Group), names_to = "Timing", values_to = "score") %>% 
  ggplot(aes(score, Group, color = Timing))
p11 <- p1 + geom_boxplot() 
p11

```

## Analysis - option 1

-   The baseline (Pre) scores appear to be similar in the two groups\
-   Therefore, we can just compare the followup scores\
-   Based on the descriptive statistics and the box plots we could argue that there is no need for a formal hypothesis test\
-   We'll just try the t test

## Comparison of followup scores using the t test

```{r}
#| comment: ""

tt1 <- t.test(Post~Group, dfw)
tt1
```

## Check the assumptions

-   Normality\

```{r}
with(dfw, shapiro.test(Post))  

```

## Check the assumptions

-   Equal variances?\

```{r}
var.test(Post ~ Group, data = dfw)
```

## Analysis - option 2

-   Calculate the reduction/ fall in the score from Pre to Post
-   Compare the two Groups

```{r}
dfw$Fall <- dfw$Pre - dfw$Post

dfw %>% dplyr::select(Group, Fall) %>% 
  tbl_summary(by = Group,
              statistic = all_continuous() ~ "{mean} ({sd})",
              digits = list(Fall ~ c(1, 1))) %>%
  as_gt() %>%
  tab_options(table.font.size = 36)
```

## Comparison of fall in score using the t test

```{r}
tt2 <- t.test(Fall~Group, dfw)
tt2
```

## Scatter plot

```{r}
p0 <- ggplot(data = dfw, aes(Pre, Post, colour = Group)) 
p1 <- p0 + geom_point() + geom_abline(intercept = 0, slope = 1)
p1


```

## ANCOVA

```{r}
r_anco <- aov(Post ~ Pre + Group, data = dfw)
summary(r_anco)
r_anco$coefficients
```

```{r include = FALSE}
c1 <- round(r_anco$coefficients[1])  
c2 <- round(r_anco$coefficients[2], 2)  
c3 <- round(r_anco$coefficients[3])

r_t <- round(cor(dfw$Pre, dfw$Post), 2)
r_I <- round(cor(dfw$Pre[dfw$Group == "I"], dfw$Post[dfw$Group == "I"]), 2)
r_C <- round(cor(dfw$Pre[dfw$Group == "C"], dfw$Post[dfw$Group == "C"]), 2)

s_anco <- summary(r_anco)
Tssq <- sum(s_anco[[1]][["Sum Sq"]])
ssq <- s_anco[[1]][["Sum Sq"]]
Rsquared <- round(1-(ssq[3]/Tssq),2)

```

## Formula

```{r}

print(noquote(paste0("Post = ", c1, " +", c2, "XPre ", c3, "XGroup-I")))  

print(paste0("Coefficient of determination = ", Rsquared))
```

Coefficient of determination - $R^2$ $$R^2 = 1 - \frac{Res\ SSq}{Total\ SSq}$$\
- Res SSQ - Residual/ Within group sum of squares\
- Total SSq - Total sum of squares

## Plot to assess Homogeniety of Variance

```{r}
plot(r_anco, 1)
```

## Bartlett test{.smaller}

```{r}
a <- as.data.frame(cbind(r_anco$residuals, r_anco$xlevels$Group))
a$V1 <- as.numeric(a$V1)


bartlett.test(V1 ~ V2, data = a)

```

## Levene test

```{r}
leveneTest(V1 ~ V2, data = a)

```

## Plot to assess Normality

```{r}
plot(r_anco, 2)
```

## Normality

Can be assessed by the Shapiro-Wilk normality test

```{r}
shapiro.test(residuals(r_anco))
```

## Scatterplot

```{r}
p1 + 
  labs(caption = paste0("r all - ", r_t, "; r Group-I - ", r_I, "; r Group_C - ", r_C)) +
  theme(plot.caption = element_text(size = 16))


```

## Method of analysis, power & sample size {.smaller}

-   ANCOVA most powerful, therefore required sample size lowest\
-   When the correlation between baseline & followup is high (r \~ 0.8) change score will have almost similar power\
-   When the correlation between baseline & followup is low (r \< 0.3) comparison of followup score is more powerful than change score provided there is no baseline imbalance
-   The method of analysis should be specified in the protocol. Trying all three to find a significant results is wrong.
-   In general for this type of data ANCOVA is the method of choice

## Factorial design

For clinical trials comparing several treatments.\  

\  

#### Trial comparing low-dose adrenaline, promethazine, and hydrocortisone in the prevention of acute adverse reactions to antivenom following snakebite

## Latin square design

|     | C1  | C2  | C3  | C4  |
|-----|-----|-----|-----|-----|
| R1  |     |     |     |     |
| R2  |     |     |     |     |
| R3  |     |     |     |     |
| R4  |     |     |     |     |

## Latin square design

|     | C1  | C2  | C3  | C4  |
|-----|-----|-----|-----|-----|
| R1  | A   | B   | C   | D   |
| R2  | B   | C   | D   | A   |
| R3  | C   | D   | A   | B   |
| R4  | D   | A   | B   | C   |

How many different ways are possible?

## Example 5x5x5 Latin square


```{r}
fertil <- c(rep("fertil1",1), rep("fertil2",1), rep("fertil3",1), rep("fertil4",1), rep("fertil5",1))
till <- c(rep("treatA",5), rep("treatB",5), rep("treatC",5), rep("treatD",5), rep("treatE",5))
seed <- c("A","E","C","B","D", "C","B","A","D","E", "B","C","D","E","A", "D","A","E","C","B", "E","D","B","A","C")
yield <- c(42,45,41,56,47, 47,54,46,52,49, 55,52,57,49,45, 51,44,47,50,54, 44,50,48,43,46)

df1 <- as_tibble(data.frame(till, fertil, seed, yield))

df1
```


## Yield by fertilizer

```{r}
df1 %>% 
  dplyr::select(fertil, yield) %>% 
  tbl_summary(by = fertil,
              statistic = all_continuous() ~ "{mean} ({sd})",
              digits = list(yield ~ c(1, 1))) %>%
  as_gt() %>%
  tab_options(table.font.size = 36)
```

## Yield by tillage

```{r}
df1 %>% 
  dplyr::select(till, yield) %>% 
  tbl_summary(by = till,
              statistic = all_continuous() ~ "{mean} ({sd})",
              digits = list(yield ~ c(1, 1))) %>%
  as_gt() %>%
  tab_options(table.font.size = 36)
```

## Yield by seed

```{r}
df1 %>% 
  dplyr::select(seed, yield) %>% 
  tbl_summary(by = seed,
              statistic = all_continuous() ~ "{mean} ({sd})",
              digits = list(yield ~ c(1, 1))) %>%
  as_gt() %>%
  tab_options(table.font.size = 36)
```

## Plot

```{r}


#| message: false
f <- ggplot(data = df1, aes(x = factor(fertil), y =yield)) +
  geom_dotplot(binaxis = "y", stackdir = "center") + 
  xlab("") + ylab("") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10))

t <- ggplot(data = df1, aes(x = factor(till), y =yield)) +
  geom_dotplot(binaxis = "y", stackdir = "center") + 
  xlab("") + ylab("") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10))

s <- ggplot(data = df1, aes(x = factor(seed), y =yield)) +
  geom_dotplot(binaxis = "y", stackdir = "center") + 
  xlab("") + ylab("") +
  theme(axis.text=element_text(size=10),
        axis.title=element_text(size=10))

plot_grid(f, t, s, nrow = 1)

```

## 3 way Anova


```{r}
a1 <- aov(yield ~ fertil + till + seed, data = df1)

tidy(anova(a1)) %>% 
  add_row(term = "Total", df = sum(.$df), sumsq = sum(round(.$sumsq))) %>% 
  knitr::kable(digits = c(0, 0, 1, 1, 2, 4))
```


